# Results
#### Required Packages and Data Loading
```{r}
# install.packages(c("dplyr","tidyr","ggplot2","ggalluvial","forcats",'ggtext'))
library(dplyr) 
library(tidyr)
library(tidyverse)
library(ggplot2)
library(ggalluvial)
library(forcats)
library(ggridges)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(patchwork)
library(ggrepel)
library(vcd)
library(scales) # Required for label_number()
#library(Lock5withR)
```

```{r}
# Loading in the PISA datasets:
pisa2009 <- read.csv('data/2009_PISA.csv')
pisa2012 <- read.csv('data/2012_PISA.csv')
pisa2015 <- read.csv('data/2015_PISA.csv')
pisa2018 <- read.csv('data/2018_PISA.csv')
pisa2022 <- read.csv('data/2022_PISA.csv')

participation_list <- list(
  '2009' = pisa2009$X,
  '2012' = pisa2012$Country,
  '2015' = pisa2015$X,
  '2018' = pisa2018$Country,
  '2022' = pisa2022$Country
)
all_countries <- unique(unlist(participation_list))
```

```{r}
# Reading Scores Full Table
all_reading_scores <- data.frame(country = all_countries) |>
  left_join(select(pisa2009, X, On.the.overall.reading.scale), by=c("country" = "X")) |>
  left_join(select(pisa2012, Country, Mean.score.in.PISA.2012...Reading), by=c("country" = "Country")) |>
  left_join(select(pisa2015, X, Mean.score.in.PISA.2015...Reading), by=c("country" = "X")) |>
  left_join(select(pisa2018, Country, Reading.Mean), by=c("country" = "Country")) |>
  left_join(select(pisa2022, Country, Reading.Mean), by=c("country" = "Country")) |>
  rename(`2009`=On.the.overall.reading.scale, `2012`=Mean.score.in.PISA.2012...Reading, `2015`=Mean.score.in.PISA.2015...Reading, `2018`=Reading.Mean.x, `2022`=Reading.Mean.y) |>
  mutate(across(c(`2009`, `2012`, `2015`, `2018`, `2022`), as.numeric))
  
all_reading_scores_long <- all_reading_scores |>
  pivot_longer(!country, names_to="year", values_to="score") |>
  mutate(across(c(year), as.numeric))

# Math Scores Full Table
all_math_scores <- data.frame(country = all_countries) |>
  left_join(select(pisa2009, X, On.the.mathematics.scale), by=c("country" = "X")) |>
  left_join(select(pisa2012, Country, Mean.score.in.PISA.2012...Mathematics), by=c("country" = "Country")) |>
  left_join(select(pisa2015, X, Mean.score.in.PISA.2015...Mathematics), by=c("country" = "X")) |>
  left_join(select(pisa2018, Country, Mathematics.Mean), by=c("country" = "Country")) |>
  left_join(select(pisa2022, Country, Mathematics.Mean), by=c("country" = "Country")) |>
  rename(`2009`=On.the.mathematics.scale, `2012`=Mean.score.in.PISA.2012...Mathematics, `2015`=Mean.score.in.PISA.2015...Mathematics, `2018`=Mathematics.Mean.x, `2022`=Mathematics.Mean.y) |>
  mutate(across(c(`2009`, `2012`, `2015`, `2018`, `2022`), as.numeric))
  
all_math_scores_long <- all_math_scores |>
  pivot_longer(!country, names_to="year", values_to="score") |>
  mutate(across(c(year), as.numeric))

# Science Scores Full Table
all_science_scores <- data.frame(country = all_countries) |>
  left_join(select(pisa2009, X, On.the.science.scale), by=c("country" = "X")) |>
  left_join(select(pisa2012, Country, Mean.score.in.PISA.2012...Science), by=c("country" = "Country")) |>
  left_join(select(pisa2015, X, Mean.score.in.PISA.2015...Science), by=c("country" = "X")) |>
  left_join(select(pisa2018, Country, Science.Mean), by=c("country" = "Country")) |>
  left_join(select(pisa2022, Country, Science.Mean), by=c("country" = "Country")) |>
  rename(`2009`=On.the.science.scale, `2012`=Mean.score.in.PISA.2012...Science, `2015`=Mean.score.in.PISA.2015...Science, `2018`=Science.Mean.x, `2022`=Science.Mean.y) |>
  mutate(across(c(`2009`, `2012`, `2015`, `2018`, `2022`), as.numeric))
  
all_science_scores_long <- all_science_scores |>
  pivot_longer(!country, names_to="year", values_to="score") |>
  mutate(across(c(year), as.numeric))

# Combining all scores
all_scores_long <- bind_rows(
  mutate(all_reading_scores_long, subject = "reading"),
  mutate(all_math_scores_long, subject = "math"),
  mutate(all_science_scores_long, subject = "science")
)
```

### global trends 
- we've already introduced PISA
- so let's look at the ridge plot 
- we say that the data is being more bimodal over time 
- we dont know what is exactly changing; we can say data from higher peak is going to lower peak but we are still unsure af 
```{r}
ggplot(all_scores_long, aes(x = score, y = as.factor(year))) +
  geom_density_ridges(fill = "blue") +
  ggtitle("Overall PISA Scores Over Time") +
  ylab("year")
```
- change the ordering of the years axis 
- data is being more bimodal over time 




- Let's look closer into these differences by country
```{r}
all_scores_long |>
  filter(year > 2017) |>
  filter(!is.na(score)) |>
  ggplot(aes(x = score, y = fct_reorder2(country, year=="2022", score), color=as.factor(year))) +
  geom_point(alpha=0.5) +
  ggtitle("PISA Scores in 2018 vs. 2022") +
  ylab("") +
  theme_linedraw()
```
- make a note about the three colors that are appearing 
- we see that actually the countries whose scores differed greatly were higher performing countries; that's wack 
- the change is getting worse 


- was the decrease from 2018 to 2022 significantly worse? Let's look at the whole distribution
```{r}
# Plotting scores
ggplot(filter(all_reading_scores_long, country == "OECD average"), aes(year, score)) + 
  geom_point(data = filter(all_reading_scores_long, year > 2010), aes(x=year, y=score), position='jitter') +
  geom_point(size=4, color="red") +
  labs(x = "Year", y = "OECD Average Reading Scores") +
  ggtitle("OECD Average Reading Scores from 2012-2022")

ggplot(filter(all_math_scores_long, country == "OECD average"), aes(x=year, y=score)) + 
  geom_point(data = filter(all_math_scores_long, year > 2010), aes(x=year, y=score), position='jitter') +
  geom_point(size=4, color="red") +
  labs(x = "Year", y = "OECD Average Math Scores") +
  ggtitle("OECD Average Math Scores from 2012-2022")

ggplot(filter(all_science_scores_long, country == "OECD average"), aes(year, score)) + 
  geom_point(data = filter(all_science_scores_long, year > 2010), aes(x=year, y=score), position='jitter') +
  geom_point(size=4, color="red") +
  labs(x = "Year", y = "OECD Average Science Scores") +
  ggtitle("OECD Average Science Scores from 2012-2022")

# Change the x axis to only show the years we wanna see
```


- Hard to tell. Let's try to predict 2022
```{r}
# Reading Predictions
reading_scores_train <- all_reading_scores_long |>
  filter(year < 2020, country == "OECD average")
reading_model <- lm(score ~ year, data = reading_scores_train)

# Predict for 2022
predicted_value <- predict(reading_model, newdata = data.frame(year = 2022))
print(predicted_value)

# Plot new model
ggplot(reading_scores_train, aes(year, score)) + 
  geom_point(data=filter(all_reading_scores_long, country == "OECD average"), aes(year, score)) +
  geom_point(data=data.frame(year = 2022, score = predicted_value), color = "blue", size=3) +
  geom_abline(intercept = coef(reading_model)[1], slope = coef(reading_model)[2], 
              color = "blue", linewidth = 1) +
  labs(x = "Year", y = "OECD Average Reading Scores") +
  ggtitle("Predicting 2022 OECD Average Reading Score")

# Math Predictions
math_scores_train <- all_math_scores_long |>
  filter(year < 2020, country == "OECD average")
math_model <- lm(score ~ year, data = math_scores_train)

# Predict for 2022
predicted_value <- predict(math_model, newdata = data.frame(year = 2022))
print(predicted_value)

# Plot new model
ggplot(math_scores_train, aes(year, score)) + 
  geom_point(data=filter(all_math_scores_long, country == "OECD average"), aes(year, score)) +
  geom_point(data=data.frame(year = 2022, score = predicted_value), color = "blue", size=3) +
  geom_abline(intercept = coef(math_model)[1], slope = coef(math_model)[2], 
              color = "blue", linewidth = 1) +
  labs(x = "Year", y = "OECD Average Mathematics Scores") +
  ggtitle("Predicting 2022 OECD Average Mathematics Score")

# Science Predictions
science_scores_train <- all_science_scores_long |>
  filter(year < 2020, country == "OECD average")
science_model <- lm(score ~ year, data = science_scores_train)

# Predict for 2022
predicted_value <- predict(science_model, newdata = data.frame(year = 2022))
print(predicted_value)

# Plot new model
ggplot(science_scores_train, aes(year, score)) + 
  geom_point(data=filter(all_science_scores_long, country == "OECD average"), aes(year, score)) +
  geom_point(data=data.frame(year = 2022, score = predicted_value), color = "blue", size=3) +
  geom_abline(intercept = coef(science_model)[1], slope = coef(science_model)[2], 
              color = "blue", linewidth = 1) +
  labs(x = "Year", y = "OECD Average Science Scores") +
  ggtitle("Predicting 2022 OECD Average Science Score")
```

- reading and math scores dropped significantly 
- science was ok somehow
- yes there were only three years so we cant trust this trend too much 
- but each point was being created by the average of millions of students across the world
- we have a lack of time(years) for this data but we dont have a lack of data


### Other factors
- intro to this section is gonna be: we have now concluded that there is a sig dip from 2018-2022 but we can't say it's just bc of covid so lets look at some other factors 
---> see if there is a correlation between COVID and GDP, COVID and spending


- pisa vs spending
  - diff countries spend diff 
```{r}

 #| fig-width: 8
 #| fig-height: 20

pisa_Spending2018 <- inner_join(
  read.csv("data/2018_GDP.csv"),
  read.csv("data/2018_Spending.csv"),
  by = "Country"
) %>%
  select(Country, Score, Spending) %>%
  mutate(
    Score = as.numeric(Score),
    Spending = as.numeric(Spending)
  )

pisa_points <- pisa_Spending2018 %>%
  filter(Country != "OECD average")

oecd_avg_score <- pisa_Spending2018 %>% 
  filter(Country == "OECD average") %>% 
  pull(Score)

mean_s <- pisa_points %>% 
  summarize(mean_s = mean(Spending, na.rm = TRUE)) %>% 
  pull(mean_s)

pisa_points <- pisa_points %>%
  mutate(
    Above_Both = Score > oecd_avg_score & Spending > mean_s
  )


# --- Improved Plot ---
ggplot(pisa_points, aes(x = Spending, y = Score)) +
  
  # Points
  geom_point(aes(color = Above_Both), size = 3.8) +
  
  # Nice, readable text labels
  geom_text_repel(
    aes(label = Country),
    size = 3.2,
    box.padding = 0.5,
    point.padding = 0.3,
    min.segment.length = 0,
    max.overlaps = Inf,
    segment.alpha = 0.3
  ) +
  
  # Reference lines
  geom_hline(yintercept = oecd_avg_score, 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean_s, 
             color = "blue", linetype = "dotted", size = 1) +

  # Custom colors
  scale_color_manual(
    values = c("FALSE" = "grey40", "TRUE" = "#1f78b4"),
    labels = c("Other countries", "Above both thresholds"),
    name = ""
  ) +
  scale_x_continuous(labels = scales::comma)+

  # Give breathing room so text isn't cut off
  coord_cartesian(
    clip = "off",
    xlim = c(min(pisa_points$Spending) * 0.95, max(pisa_points$Spending) * 1.05),
    ylim = c(min(pisa_points$Score) * 0.97, max(pisa_points$Score) * 1.03)
  ) +

  labs(
    title = "PISA 2018 Reading Score vs Spending",
    subtitle = "Countries above OECD average reading score AND above Spending are highlighted",
    x = "Spending",
    y = "PISA Reading Score"
  ) +

  # Much cleaner theme
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom",

    # Remove excessive gridlines
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "grey85"),
    panel.grid.major.x = element_line(color = "grey90"),

    # Prevent clipping of labels
    plot.margin = margin(20, 20, 20, 20)
  )

```

```{r}

 #| fig-width: 8
 #| fig-height: 12

pisa_Spending2022 <- inner_join(
  read.csv("data/2022_Spending.csv"),
  read.csv("data/2022_PISA.csv"),
  by = "Country"
) %>%
  select(Country, Reading.Mean, Spending) %>%
  mutate(
    Reading.Mean = as.numeric(Reading.Mean),
    Spending = as.numeric(Spending)
  )

pisa_points <- pisa_Spending2022 %>%
  filter(Country != "OECD average")

oecd_avg_score <- pisa_Spending2022 %>% 
  filter(Country == "OECD average") %>% 
  pull(Reading.Mean)

mean_sp <- pisa_points %>% 
  summarize(mean_sp = mean(Spending, na.rm = TRUE)) %>% 
  pull(mean_sp)

pisa_points <- pisa_points %>%
  mutate(
    Above_Both = Reading.Mean > oecd_avg_score & Spending > mean_sp
  )

# --- Improved Plot ---
ggplot(pisa_points, aes(x = Spending, y = Reading.Mean)) +
  
  # Points
  geom_point(aes(color = Above_Both), size = 3.8) +
  
  # Nice, readable text labels
  geom_text_repel(
    aes(label = Country),
    size = 3.2,
    box.padding = 0.5,
    point.padding = 0.3,
    min.segment.length = 0,
    max.overlaps = Inf,
    segment.alpha = 0.3
  ) +
  
  # Reference lines
  geom_hline(yintercept = oecd_avg_score, 
             color = "red", linetype = "dashed", size = 1) +
  geom_vline(xintercept = mean_sp, 
             color = "blue", linetype = "dotted", size = 1) +

  # Custom colors
  scale_color_manual(
    values = c("FALSE" = "grey40", "TRUE" = "#1f78b4"),
    labels = c("Other countries", "Above both thresholds"),
    name = ""
  ) +

  # Give breathing room so text isn't cut off
  coord_cartesian(
    clip = "off",
    xlim = c(min(pisa_points$Spending) * 0.95, max(pisa_points$Spending) * 1.05),
    ylim = c(min(pisa_points$Reading.Mean) * 0.97, max(pisa_points$Reading.Mean) * 1.03)
  ) +

  labs(
    title = "PISA 2022 Reading Score vs Spending",
    subtitle = "Countries above OECD average reading score AND above mean Spending",
    x = "Spending",
    y = "PISA Reading Score"
  ) +

  # Much cleaner theme
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 12),
    legend.position = "bottom",

    # Remove excessive gridlines
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "grey85"),
    panel.grid.major.x = element_line(color = "grey90"),

    # Prevent clipping of labels
    plot.margin = margin(20, 20, 20, 20)
  )

```

- graph: spendin 2018 - 2022
  - woah change or no change 
    - change could be bc of gdp if a countru has more money they will spend more on ending 

GRAPH TO BE CREATED

- GDP vs spending 
  - there is a corr 
```{r}
GDP_precovid<- read.csv('data/2018_GDP.csv')
spending_precovid<- read.csv('data/2018_Spending.csv')

merged_data <- merge(GDP_precovid, spending_precovid, by = "Country")
```

```{r}

#2018

# Define the specific countries you want to label
target_countries <- c("Qatar", "Macao (China)", "Singapore", "Austria", "Iceland",'Finland')

ggplot(merged_data, aes(x = GDP, y = Spending)) +
  geom_point(color = "blue", size = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  
  geom_text_repel(
    # Filter data to only label the target countries
    data = merged_data[merged_data$Country %in% target_countries, ], 
    aes(label = Country), 
    size = 3,                       
    segment.color = 'grey50',       
    segment.size = 0.2,             
    box.padding = unit(0.5, "lines"), 
    point.padding = unit(0.5, "lines"),
    force = 2
  ) +
  
  # Set X-axis limits and format (NO scientific notation)
  scale_x_continuous(
    labels = label_number(big.mark = ","), # Format to full number with commas
    limits = c(0, max(merged_data$GDP) * 1.05) # Incorporate the limit here
  ) +
  
  # Set Y-axis limits and format (NO scientific notation)
  scale_y_continuous(
    labels = label_number(big.mark = ","), # Format to full number with commas
    limits = c(0, max(merged_data$Spending) * 1.05) # Incorporate the limit here
  ) +
  
  # Labels and Theme
  labs(
    title = "GDP vs. Spending Correlation (2018) - Selected Countries Highlighted",
    x = "GDP (in currency units from table)",
    y = "Spending (in currency units from table)",
    caption = "Linear trend line (dark red) shows relationship. Dashed line (grey) is y=x."
  ) +
  theme_minimal() 

# dashed is y = x
```

```{r}

GDP_postcovid<- read.csv('data/2022_GDP.csv')
spending_postcovid<- read.csv('data/2022_Spending.csv')

merged_data2 <- merge(GDP_postcovid, spending_postcovid, by = "Country")
names(merged_data2)[2] <- "GDP"
```

```{r}

# Define the specific countries you want to label
target_countries <- c("Qatar", "Macao (China)", "Luxembourg", "Singapore", "Austria", "Iceland", 'Finland')

ggplot(merged_data2, aes(x = GDP, y = Spending)) +
  geom_point(color = "blue", size = 2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  
  geom_text_repel(
    # Filter data to only label the target countries
    data = merged_data2[merged_data2$Country %in% target_countries, ], 
    aes(label = Country), 
    size = 3,                       
    segment.color = 'grey50',       
    segment.size = 0.2,             
    box.padding = unit(0.5, "lines"), 
    point.padding = unit(0.5, "lines"),
    force = 2
  ) +
  
  # Set X-axis limits and format (NO scientific notation)
  scale_x_continuous(
    labels = label_number(big.mark = ","), # Format to full number with commas
    limits = c(0, max(merged_data2$GDP) * 1.05) # Incorporate the limit here
  ) +
  
  # Set Y-axis limits and format (NO scientific notation)
  scale_y_continuous(
    labels = label_number(big.mark = ","), # Format to full number with commas
    limits = c(0, max(merged_data2$Spending) * 1.05) # Incorporate the limit here
  ) +
  
  # Labels and Theme
  labs(
    title = "GDP vs. Spending Correlation (2022) - Selected Countries Highlighted",
    x = "GDP (in currency units from table)",
    y = "Spending (in currency units from table)",
    caption = "Linear trend line (dark red) shows relationship. Dashed line (grey) is y=x."
  ) +
  theme_minimal() 

# dashed is y = x
```

- pisa vs gdp 
  -bc gdp and spending are correlated we see the same type of corr so this prove that gdp and spending are correlated 
- GRAPH TO BE CREATED
  
we end on how has GDP changed (graph: GDP 2018-2022)
- get rid of insane colors 
- color with red and green--> threshold of the expected slope 
```{r}
gdp_2018 <- read.csv('data/2018_GDP.csv') |>
  mutate(country = Country, score = Score, gdp = GDP, year="2018") |>
  select("country", "score", "gdp", "year")
gdp_2022 <- read.csv('data/2022_GDP.csv') |>
  left_join(filter(all_reading_scores_long, year == 2022), by=c("Country" = "country")) |>
  mutate(country = Country, gdp = GDP, year="2022") |>
  select("country", "score", "gdp", "year")

gdp <- bind_rows(gdp_2018, gdp_2022) 
# |>
  #pivot_wider(names_from = year, values_from = gdp, names_prefix = "gdp_", values_fn = first) |>
  #mutate(change = gdp_2022 - gdp_2018)

ggplot(gdp, aes(x = as.factor(year), y = gdp, group = country)) +
  geom_line(aes(color = country), linewidth = 1) +
  geom_point(size = 3) +
  labs(title = "GDP Trend: 2018 to 2022",
       x = "Year", y = "GDP") +
  theme(legend.position = "none") 
```
- has not increased as much as it did previously
- GDP will typically increase but the increase that we see here is not as high as the increase that we would typically want to see - largest reason for that is covid 
- though we can't say that COVID is the only reason why the education system has changed, a lot of the other factors that affect education outcomes were effected by covid 



### CASE STUDY: FINLAND
Finland's transition to distance learning went well, but the effect of switching meant that students no longer had equal access to the resources provided at school and their resources were limited to their homes. Meaning, any difference or inequity of educational support in their home life directly affected their educational performance/outcomes. This can be seen in their cynicism and burnout or whatever rates increasing
```{r}
data_cynicism <- read.csv("data/cynicism_finland.csv")
data_cynicism <- data_cynicism %>%
  rename(
    High_school = colnames(data_cynicism)[4],
  )

# Convert data to long format for easier plotting
data_cynicism_long <- data_cynicism %>%
  pivot_longer(
    # Select the three school columns for pivoting
    cols = c(High_school),
    names_to = "School_Type",
    values_to = "Percentage"
  )


color_map <- c(
  "High_school" = "#6495ED"  
)

label_map <- c(
  "Comprehensive" = "Comprehensive school grades 8 and 9",
  "High_school" = "High school grades 10 and 11",
  "Vocational" = "Vocational school"
)

# Create the plot
cynicism_plot <- ggplot(data_cynicism_long, aes(x = Year, y = Percentage, color = School_Type)) +
  # Add the lines
  geom_line(linewidth = 1) +
  # Add the points (circles)
  geom_point(size = 3) +
  
  # Set the custom colors and labels for the legend
  scale_color_manual(
    values = color_map, 
    labels = label_map,
    name = NULL # Remove legend title
  ) +
  
  # Customize the X-axis (Years)
  scale_x_continuous(
    breaks = data_cynicism$Year,
    # Use the custom labels, noting 2013 and 2017 are single years in this set
    labels = c("2007", "2009", "2011", "2013", "2017", "2019", "2021"),
    expand = expansion(add = c(0.5, 0.5))
  ) +
  
  # Customize the Y-axis (Percentage) - Adjusted for higher values
  scale_y_continuous(
    limits = c(0, 40), # Set the y-axis range to accommodate up to 35.5
    breaks = seq(0, 40, by = 5) # Set breaks every 5 units
  ) +
  
  # Add labels
  labs(
    x = "Year",
    y = "Percentage of students reporting cynicism"
  ) +
  
  # Apply the clean theme
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.title = element_text(size = 12, face = "bold"),
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_blank()
  )
# Display the plot
print(cynicism_plot)

# 2020 vertical line to signify the start of pandemic 
# some kinda predictive thing to show that predicted value for 2021 is lower than actual to show hella increase in slope
```

```{r}
data <- read.csv("data/school_burnout_finland.csv")
data_long <- data %>%
  rename(
    High_school = colnames(data)[4]
  )
```


```{r}
# Ensure Year is numeric
data_long$Year <- as.numeric(data_long$Year)

# --- Regression and Prediction Setup ---
# 2. Prepare the training data (up to 2019)
data_train <- subset(data_long, Year <= 2019)

# 3. Fit the Linear Regression Model
burnout_model <- lm(High_school ~ Year, data = data_train)

# 4. Predict the High_school value for 2021
prediction_year <- data.frame(Year = 2021)
predicted_value <- predict(burnout_model, newdata = prediction_year)

# 5. Create data frames for plotting the actual and predicted points
predicted_point <- data.frame(
  Year = 2021, 
  High_school = predicted_value,
  Type = "Predicted"
)

# Create a data frame for all actual points, distinguishing the 2021 point
actual_points <- data_long
actual_points$Type <- "Actual (2009-2019)"
actual_points[actual_points$Year == 2021, "Type"] <- "Actual (2021)"

# --- Combined Plot Generation ---

ggplot(data = actual_points, aes(x = Year, y = High_school)) +
  
  # A. Add the vertical shaded section (starting from 2020)
  annotate(
    geom = "rect",
    xmin = 2020, 
    xmax = 2022, # Extends past 2021 to cover the right side of the plot
    ymin = -Inf, 
    ymax = Inf,
    fill = "grey", 
    alpha = 0.3
  ) +
  
  # B. Plot the regression line (fitted on 2009-2019 data)
  geom_smooth(
    data = data_train, # Base the line on the training data
    method = "lm", 
    se = FALSE, # Do not display standard error
    color = "darkgreen", 
    linetype = "dashed", 
    linewidth = 1
  ) +
  
  # C. Plot all actual data points
  geom_point(aes(color = Type), size = 3) + 
  
  # D. Highlight the predicted 2021 point
  geom_point(
    data = predicted_point, 
    aes(color = Type), 
    size = 4, 
    shape = 17 # Triangle shape
  ) +
  
  # E. Connect the last actual point (2019) to the predicted 2021 point
  geom_segment(
    aes(
      x = 2019, 
      y = actual_points[actual_points$Year == 2019, "High_school"], 
      xend = 2021, 
      yend = predicted_point[predicted_point$Year == 2021, "High_school"]
    ),
    linetype = "dotted",
    color = "darkgreen",
    linewidth = 1
  ) +
  
  # 👇 This forces the y-axis to include 0
  expand_limits(y = 0) + 
  
  # Add labels and theme...
  labs(
    title = "Student Burnout (Persistent Sadness) Trends: Finland",
    subtitle = paste0(
      "Linear Prediction for 2021: ", round(predicted_value, 2), 
      "% (Actual: ", actual_points[actual_points$Year == 2021, "High_school"], "%)"
    ),
    y = "Percent of Students (%)",
    x = "Year"
  ) +
  theme_minimal() +
  
  # F. Customize colors and legend
  scale_color_manual(
    name = "Data Point",
    values = c(
      "Actual (2009-2019)" = "black", 
      "Actual (2021)" = "red", 
      "Predicted" = "blue"
    )
  ) +
  
  # Move the legend to a convenient location
  theme(legend.position = "bottom")

# 2020 vertical line to signify the start of pandemic 
# some kinda predictive thing to show that predicted value for 2021 is lower than actual to show hella increase in slope
```
- graph shows spike from 2019-2021


The transition to distance learning during the pandemic led to decreased engagement among students, teachers, and principals. This shift significantly increased stress and burnout for educators, with principals feeling the effects of their teachers' struggles. Teachers found it difficult to compensate for the inconsistent support students received back home. As seen in this teacher burnout graph:
GRAPH OF TEACHER BURNOUT TO BE CREATED


Conclusion: The switch to distance learning was well-organized, but it ultimately made fairness in education worse and damaged the conditions needed for students to learn well and feel supported.








### CASE STUDY: USA 
- find a source to talk about whether the transition from diff modalities is trash and just generally how that transition went
```{r}

#### school modality 

#https://healthdata.gov/browse?q=school+learning+modalities&sortBy=relevance&pageSize=20

df <- read.csv("data/school_modality.csv") |>
  mutate(
    Learning_Modality = as.factor(Learning_Modality),
    Year = as.factor(Year)
  )

counts <- df |>
  group_by(Year, Learning_Modality) |>
  summarize(Freq = sum(Count), .groups = "drop")

tbl <- xtabs(Freq ~ Year + Learning_Modality, data = counts)

colors_theme <- c("#FFB6B9", "#FAE3D9", "#61C0BF")

mosaic(
  tbl,
  direction = c("v", "h"),
  highlighting = "Learning_Modality",
  highlighting_fill = colors_theme,
  main = "Learning Modality by Year")
```
- from 2020 to 2022, the USA changed learning modality due to pandemic
- by the end of 2022, schools were back to hybrid
- "K-12 public and independent charter school districts for the 2021-2022 school year and the Fall 2022 semester, from August 2021 – December 2022"

```{r}
# HIGH SCHOOL STUDENTS
usaburnout <- read.csv("data/USA_burnout.csv")
# Ensure Year is numeric
usaburnout$Year <- as.numeric(usaburnout$Year)

# --- Regression and Prediction Setup ---

# 2. Prepare the training data (up to 2019)
usaburnout_train <- subset(usaburnout, Year <= 2019)

# 3. Fit the Linear Regression Model (using data up to 2019)
burnout_model <- lm(High_School ~ Year, data = usaburnout_train)

# 4. Predict the High_School value for 2021
prediction_year <- data.frame(Year = 2021)
predicted_value <- predict(burnout_model, newdata = prediction_year)

# 5. Create data frames for plotting the actual and predicted points
predicted_point <- data.frame(
  Year = 2021, 
  High_School = predicted_value,
  Type = "Predicted"
)

# Create a data frame for all actual points, distinguishing the 2021 point
actual_points <- usaburnout
actual_points$Type <- "Actual (2007-2019)"
actual_points[actual_points$Year == 2021, "Type"] <- "Actual (2021)"

# --- Combined Plot Generation ---

ggplot(data = actual_points, aes(x = Year, y = High_School)) +
  
  # A. Add the vertical shaded section (starting from 2020)
  annotate(
    geom = "rect",
    xmin = 2020, 
    xmax = 2022, # Extends past 2021 to cover the right side of the plot
    ymin = -Inf, 
    ymax = Inf,
    fill = "grey", 
    alpha = 0.3
  ) +
  
  # B. Plot the regression line (fitted on 2007-2019 data)
  geom_smooth(
    data = usaburnout_train, # Base the line on the training data
    method = "lm", 
    se = FALSE, # Do not display standard error
    color = "darkgreen", 
    linetype = "dashed", 
    linewidth = 1
  ) +
  
  # C. Plot all actual data points
  geom_point(aes(color = Type), size = 3) + 
  
  # D. Highlight the predicted 2021 point
  geom_point(
    data = predicted_point, 
    aes(color = Type), 
    size = 4, 
    shape = 17 # Triangle shape
  ) +
  
  # E. Connect the last actual point (2019) to the predicted 2021 point
  geom_segment(
    aes(
      x = 2019, 
      y = actual_points[actual_points$Year == 2019, "High_School"], 
      xend = 2021, 
      yend = predicted_point[predicted_point$Year == 2021, "High_School"]
    ),
    linetype = "dotted",
    color = "darkgreen",
    linewidth = 1
  ) +
  
  expand_limits(y = 0) + 
  
  labs(
    title = "Student Burnout (Persistent Sadness) Trends: USA",
    subtitle = paste0(
      "Linear Prediction for 2021: ", round(predicted_value, 2), 
      "% (Actual: ", actual_points[actual_points$Year == 2021, "High_School"], "%)"
    ),
    y = "Percent of Students (%)",
    x = "Year"
  ) +
  theme_minimal() +
  
  # F. Customize colors and legend
  scale_color_manual(
    name = "Data Point",
    values = c(
      "Actual (2007-2019)" = "black", 
      "Actual (2021)" = "red", 
      "Predicted" = "blue"
    )
  ) +
  
  # Move the legend to a convenient location
  theme(legend.position = "bottom")
# 2020 vertical line to signify the start of pandemic 
# some kinda predictive thing to show that predicted value for 2021 is lower than actual to show hella increase in slope

```
- check the steady increase from 2019 to 2021 (COVID TIME)
- could be because of the stress of pandemic and changes in school modality (refer to mosaic)















### Chloropleth, we are not sure if we are adding/using

```{r}
pisa_gdp2018 = inner_join(read.csv("data/2018_GDP.csv"), read.csv("data/2018_PISA.csv"), by = "Country") %>%
  select(Country, Score, GDP)

name_fix <- c(
  "United States" = "United States of America",
  "South Korea" = "Korea, Republic of",
  "North Korea" = "Korea, Democratic People's Republic of",
  "Iran" = "Iran, Islamic Republic of",
  "Vietnam" = "Viet Nam",
  "Egypt" = "Egypt, Arab Republic of",
  "Czech Republic" = "Czechia",
  "Syria" = "Syrian Arab Republic",
  "Bolivia" = "Bolivia (Plurinational State of)",
  "Brunei Darussalam" = "Brunei",
  "Slovak Republic" = "Slovakia",
  "Korea" = "South Korea",
  "Baku (Azerbaijan)" = "Azerbaijan",
  "Moldova" = "Moldova",       
  "Kosovo" = "Kosovo")

pisa_gdp2018$Country <- recode(pisa_gdp2018$Country, !!!name_fix)

world <- ne_countries(scale = "medium", returnclass = "sf")

world_data <- world %>%
  left_join(pisa_gdp2018, by = c("name" = "Country"))

map_gdp <- ggplot(world_data) +
  geom_sf(aes(fill = GDP), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "gray90") +
  labs(
    title = "GDP by Country",
    fill = "GDP"
  ) +
  theme_minimal()

map_score <- ggplot(world_data) +
  geom_sf(aes(fill = Score), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "gray90") +
  labs(
    title = "PISA Score by Country",
    fill = "Score"
  ) +
  theme_minimal()
map_gdp

map_score

```


```{r}
pisa_gdp2022 = inner_join(read.csv("data/2022_GDP.csv"), read.csv("data/2022_PISA.csv"), by = "Country") %>%
  select(Country, Reading.Mean, GDP)



name_fix <- c(
  "United States" = "United States of America",
  "South Korea" = "Korea, Republic of",
  "North Korea" = "Korea, Democratic People's Republic of",
  "Iran" = "Iran, Islamic Republic of",
  "Vietnam" = "Viet Nam",
  "Egypt" = "Egypt, Arab Republic of",
  "Czech Republic" = "Czechia",
  "Syria" = "Syrian Arab Republic",
  "Bolivia" = "Bolivia (Plurinational State of)",
  "Brunei Darussalam" = "Brunei",
  "Slovak Republic" = "Slovakia",
  "Korea" = "South Korea",
  "Baku (Azerbaijan)" = "Azerbaijan",
  "Moldova" = "Moldova",       
  "Kosovo" = "Kosovo")

pisa_gdp2022$Country <- recode(pisa_gdp2022$Country, !!!name_fix)

world <- ne_countries(scale = "medium", returnclass = "sf")

world_data <- world %>%
  left_join(pisa_gdp2022, by = c("name" = "Country"))

map_gdp <- ggplot(world_data) +
  geom_sf(aes(fill = GDP), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "gray90") +
  labs(
    title = "GDP by Country",
    fill = "GDP"
  ) +
  theme_minimal()

map_score <- ggplot(world_data) +
  geom_sf(aes(fill = Reading.Mean), color = "white", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", na.value = "gray90") +
  labs(
    title = "PISA Score by Country",
    fill = "Score"
  ) +
  theme_minimal()
map_gdp
#map_score
```


